\chapter{Winding Numbers, Indices, and Topological Insulators}
\label{chap:winding_numbers}

Physics is often concerned with quantities that can change continuously.
We measure fields, positions, and momenta, all of which can vary smoothly.
Yet, some of the most profound and robust phenomena in nature are governed by properties that are fundamentally discrete and unchangeable.
An electron has a charge of exactly $-e$, not $-0.99e$.
The magnetic flux through a superconducting ring is quantized in integer multiples of a fundamental constant.
These phenomena are not accidents of dynamics; they are consequences of topology.

Topology is the branch of mathematics concerned with properties of spaces that are preserved under continuous deformations.
From a topological point of view, a coffee mug and a donut are the same object because one can be smoothly reshaped into the other without tearing or gluing.
The property they share is the presence of a single hole.
The number of holes, an integer, is a \textit{topological invariant}.
It cannot be changed by any smooth transformation.
This chapter begins our exploration of the deep connection between the topological invariants of abstract spaces and the quantized, robustly protected properties of physical systems.
Our first and most fundamental example of such an invariant is the winding number.
 
\input{chapters/chapter01/sec1_winding_number.tex} 
\input{chapters/chapter01/sec2_fredholm_index.tex}
\input{chapters/chapter01/sec3_first_index_theorem.tex}
\input{chapters/chapter01/sec4_ssh_model.tex}
\input{chapters/chapter01/sec5_thouless_pumping.tex}

